---
title: "Tutorial"
author: "Gonzalo E. Pinilla-Buitrago"
date: "2025-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, progress = FALSE)
```

# `fastbioclim`Package: *Efficient Derivation of Bioclimatic Variables*

The `fastbioclim` package is designed to efficiently generate bioclimatic variables using two distinct *workflows*.

**In-Memory ("terra"):** The first method is based on the `terra` package and is ideal when rasters can be processed entirely in the computer's RAM.

**Out-of-Core ("tiled"):** The second method is designed for large rasters. It divides the area of interest into tiles (or grids) that are processed independently using the `exactextractr` and `Rfast` packages. This out-of-core approach allows for the analysis of data of any size, regardless of the available RAM.

The main advantage of `fastbioclim` is that it can intelligently select the most appropriate method with the argument `method = "auto"`, always ensuring the best balance between speed and memory usage.

In addition to its performance, `fastbioclim` offers great flexibility:
- It allows for the calculation of a subset of variables without needing to generate the complete set.
- It expands the set to 35 bioclimatic variables, including solar radiation (bios 20-27) and moisture summaries (bios 28-35) based on the ANUCLIM 6.1 nomenclature (Xu & Hutchinson, 2012).
- It offers the option to define custom time periods (e.g., weeks, semesters) for period-based variables (like bio08 or bio18).
- It allows the use of a real average temperature raster (parameter `tavg`), instead of the standard approximation of (tmax + tmin) / 2.
- It allows the analysis of any temporal variable (wind speed, humidity, etc.) with the same powerful and scalable architecture, using the `derive_statistics()` function.
- It allows the use of static indices for advanced control, ideal for time-series analysis (e.g., ensuring the "warmest period" always refers to the same months each year).

The functionality of fastbioclim is inspired by the `biovars()` function from the `dismo` package, with the goal of streamlining and scaling the process of creating bioclimatic variables for ecological and environmental modeling.

**Disclaimer: This Package is Under Development**

This R package is currently under development and may contain errors, bugs, or incomplete features.

Contributions and bug reports are welcome. If you encounter issues or have suggestions for improvement, please open an issue on the [GitHub](https://github.com/gepinillab/fastbioclim) repository.

## Installation

To install `fastbioclim`, you can use the `remotes` package. If you do not have it installed, you can do so by running:

```{r message=FALSE}
install.packages("remotes")
remotes::install_github("gepinillab/fastbioclim")
# Install to get the package example data 
remotes::install_github("gepinillab/egdata.fastbioclim")
```

Install and load the necessary packages:

```{r message=FALSE}
# Load libraries and install them if necessary
if (!require("terra")) {
  install.packages("terra") 
}
if (!require("future.apply")) {
  install.packages("future.apply") 
}
if (!require("progressr")) {
  install.packages("progressr") 
}
if (!require("fastbioclim")) {
  remotes::install_github("gepinillab/fastbioclim") 
}
if (!require("egdata.fastbioclim")) {
  remotes::install_github("gepinillab/egdata.fastbioclim") 
}
```

## Getting the 19 Bioclimatic Variables for Ecuador

Similar to `biovars()`, this package requires the user to provide the average climatic variables per time unit for the calculation. Traditionally, these time units correspond to monthly averages of temperature and precipitation over decades. For this example, we will use variables obtained and processed from CHELSA v2.1 (Karger et al., 2017) for Ecuador, which are available within the data package (`egdata.fastbioclim` on GitHub only).

```{r message=FALSE}
# Get a list of rasters and create a SpatRaster for each variable
# Minimum temperature
tmin_ecu <- system.file("extdata/ecuador/", package = "egdata.fastbioclim") |>
  list.files("tmin", full.names = TRUE) |> rast()
# Maximum temperature
tmax_ecu <- system.file("extdata/ecuador/", package = "egdata.fastbioclim") |>
  list.files("tmax", full.names = TRUE) |> rast()
# Precipitation
prcp_ecu <- system.file("extdata/ecuador/", package = "egdata.fastbioclim") |>
  list.files("prcp", full.names = TRUE) |> rast()

# Define the directory where the rasters will be saved
output_dir_bioclim <- file.path(tempdir(), "bioclim_ecuador")

# Get the 19 variables for Ecuador
bioclim_ecu <- derive_bioclim(
  bios = 1:19,
  tmin = tmin_ecu,
  tmax = tmax_ecu,
  prcp = prcp_ecu,
  output_dir = output_dir_bioclim,
  overwrite = TRUE
)
# Plot bio01 and bio12
plot(bioclim_ecu[[c("bio01", "bio12")]])
```

### Using Average Temperature as Input

The `fastbioclim` package also offers the option to use average temperature (defined with the `tavg` parameter) for calculating bioclimatic variables.

```{r message=FALSE}
# Average temperature
tavg_ecu <- system.file("extdata/ecuador/", package = "egdata.fastbioclim") |>
  list.files("tavg", full.names = TRUE) |> rast()
# Define the directory where the rasters will be saved
output_dir_bioclim_v2 <- file.path(tempdir(), "bioclim_ecuador_v2")

bioclim_ecu_v2 <- derive_bioclim(
  bios = 1:19,
  tavg = tavg_ecu,
  tmin = tmin_ecu,
  tmax = tmax_ecu,
  prcp = prcp_ecu,
  output_dir = output_dir_bioclim_v2,
  overwrite = TRUE
)
# Difference between bio01s when tavg is used
plot(bioclim_ecu_v2[["bio01"]] - bioclim_ecu[["bio01"]])
```

### Selecting a Subset of Variables

Often, it is not necessary to use all bioclimatic variables in our analyses. For this reason, and unlike `biovars()`, you can define in the `bios` parameter the number that identifies each of the bioclimatic variables. This way, it is not necessary to obtain all 19 variables to then select only the variables of interest. In the following example, only four variables will be obtained (bio05, bio06, bio13, and bio14). This example is somewhat faster, as it is not necessary to internally calculate the warmest/coldest or driest/wettest quarters.

```{r message=FALSE}
bios4_ecu <- derive_bioclim(
  tmin = tmin_ecu, 
  tmax = tmax_ecu, 
  prcp = prcp_ecu,
  bios = c(5, 6, 13, 14),
  overwrite = TRUE
)
plot(bios4_ecu)
```

### Building Summaries with Other Variables

Another important functionality of `fastbioclim` is the option to obtain statistics similar to bioclimatic variables but with other variables. As an example, we will create summaries of average monthly wind variables. For the quarterly interactive variables, we will use the wettest and driest quarters.

```{r message=FALSE}
wind_ecu <- system.file("extdata/ecuador/", package = "egdata.fastbioclim") |>
  list.files("wind", full.names = TRUE) |> rast()
wind_dir_ecu <- file.path(tempdir(), "wind_ecuador")

ecu_stats <- derive_statistics(
  variable = wind_ecu,
  stats = c("mean", "max", "min", "stdev", "max_period", "min_period"),
  inter_variable = prcp_ecu,
  inter_stats = c("max_inter", "min_inter"),
  prefix_variable = "wind",
  suffix_inter_max = "wettest",
  suffix_inter_min = "driest",
  overwrite = TRUE,
  output_dir = wind_dir_ecu
)
plot(ecu_stats)
```

Subsets of variables can also be constructed. In this case, we will create wind variables, but only based on the interaction with temperature, which correspond to "Wind in the warmest quarter" and "Wind in the coldest quarter".

```{r message=FALSE}
ecu_stats_v2 <- derive_statistics(
  variable = wind_ecu,
  stats = NULL,
  inter_variable = tavg_ecu,
  inter_stats = c("max_inter", "min_inter"),
  prefix_variable = "wind",
  suffix_inter_max = "warmest",
  suffix_inter_min = "coldest",
  overwrite = TRUE,
  output_dir = wind_dir_ecu
)
plot(ecu_stats_v2)
```

## Building for the Neotropics: 35 Variables

Based on the ANUCLIM 6.1 nomenclature (Xu & Hutchinson, 2012), `derive_bioclim()` also offers the option to create bioclimatic variables based on moisture and solar radiation indices. In this case, we will construct the 35 bioclimatic variables for an extent covering the Neotropics.

For this case, the "auto" method should use the "tiled" method for creating the variables. But you can also force it to use this method with the parameter `method="tiled"`. This method will divide the area of interest into tiles using the decimal degrees defined in the `tile_degrees` parameter (5 is the default value).

**Parallelization:** It is also important to mention that the 'tiled' method can be parallelized using `future::plan()`. For more information, consult the documentation of that package.

**Progress Bar:** A progress bar is available using the `progressr` package. To activate it, you need to use the function `progressr::handlers()` or `progressr::with_progress()`. For more information, consult the documentation of that package.

```{r message=FALSE}
# Get a list of rasters and create a SpatRaster for each variable
# Average temperature
tavg_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("tavg", full.names = TRUE) |> rast()
# Minimum temperature
tmin_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("tmin", full.names = TRUE) |> rast()
# Maximum temperature
tmax_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("tmax", full.names = TRUE) |> rast()
# Precipitation
prcp_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("prcp", full.names = TRUE) |> rast()
# Solar radiation
srad_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("srad", full.names = TRUE) |> rast()
# Climatic moisture index
mois_neo <- system.file("extdata/neotropics/", package = "egdata.fastbioclim") |>
  list.files("cmi", full.names = TRUE) |> rast()

# Define the directory where the rasters will be saved
output_dir_neo <- file.path(tempdir(), "bioclim_neotropics")

# Activate progress bar
# progressr::handlers(global = TRUE)

# Define parallelization plan
# future::plan("multisession", workers = 4)

# Get the 35 variables for the Neotropics
bioclim_neo <- derive_bioclim(
  bios = 1:35,
  tavg = tavg_neo,
  tmin = tmin_neo,
  tmax = tmax_neo,
  prcp = prcp_neo,
  srad = srad_neo,
  mois = mois_neo,
  method = "tiled",
  tile_degrees = 20,
  output_dir = output_dir_neo,
  overwrite = TRUE
)
print(bioclim_neo)
```

## Example with a User-Defined Region

Another useful parameter in the fastbioclim `package` is the option to provide an 'sf' object to delimit and mask an area of interest. The calculation of the bioclimatic variables will only be performed in this area.

```{r message=FALSE}
# Get areas of interest
mex <- qs2::qs_read(system.file("extdata/mex.qs2", package = "egdata.fastbioclim"))
# Get only bio10
bio10_mex <- derive_bioclim(
  bios = 10,
  tmax = tmax_neo,
  tmin = tmin_neo,
  user_region = mex,
  overwrite = TRUE,
  output_dir = file.path(tempdir(), "bio10_mex")
)
plot(bio10_mex)
```

## Example with a Static Variable

An advanced option within the package is the ability to determine static variables for the maximum and minimum variables by months (or units) and periods. This can be useful if the research questions are related to a specific time (e.g., seasons of the year) or in the construction of time series.

In this case, we will create the bio10 variable in Mexico again, but using the quarter of June, July, and August as a reference. To do this, we must create a SpatRaster that defines the period of interest. In `fastbioclim`, the reference is always the first month of the period or the unit. In this case, that month corresponds to the number 6. Therefore, we will create a raster filled with this number, to then be used as a reference in the creation of the 'bio10' variable.

```{r message=FALSE}
# Create a raster of 6s for the Neotropics
warmest <- tavg_neo[[1]]
warmest[!is.na(warmest)] <- 6
names(warmest) <- "warmest_period"
# Important: For the 'tiled' method, rasters must be saved to disk
terra::writeRaster(warmest, file.path(tempdir(), "warmest_static.tif"), overwrite = TRUE)
warmest <- terra::rast(file.path(tempdir(), "warmest_static.tif"))
plot(warmest)
```

```{r message=FALSE}
# Get only bio10
bio10_war <- derive_bioclim(
  bios = 10,
  tmax = tmax_neo,
  tmin = tmin_neo,
  user_region = mex,
  warmest_period = warmest,
  overwrite = TRUE,
  output_dir = file.path(tempdir(), "bio10_war")
)
# Differences in bio10 for Mexico
plot(bio10_mex - bio10_war)
```